The Ranked Nearest Neighbour Interchange space

### 2019-06-27 paper.tex
rm some definitions that are too standard

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is induced by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the parent of $x$ and will be denoted by $\parent(x)$.
We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space} or just $\rnni$.

@@

The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
Therefore, further notations need to be introduced.
If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

@@

>>>>

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|\rank(v) - \rank(u)| = l$.
Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
We also say that $v$ \emph{induces} the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is \emph{induced} by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the parent of $x$ and will be denoted by $\parent(x)$.
We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
A swap of the ranks of two internal nodes ($x$ and $y$) of subsequent ranks (|\rank(x) - \rank(y)| = 1) is called a \emph{rank move}.
We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or \emph{adjacent in} $\rnni$.
Note that thus defined $\rnni$ graph is a metric space with the standard shortest-path distance.

@@

The following definitions are less standard and required for the results obtained in this paper.
If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called a \emph{caterpillar tree}.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

@@

This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
